Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-16 src/article.tex
Revise manuscript for readability again

% Scaffolder url
\urldef{\scaffolder}\url{http://www.michaelbarton.me.uk/projects/scaffolder/}

@@

\author{
  Michael D Barton$^{1}$\correspondingauthor%
  \email{Michael D Barton\correspondingauthor  - mail@michaelbarton.me.uk}%
\and
  Hazel A Barton$^1$%
  \email{Hazel A Barton - bartonh@nku.edu}%
      }

@@

  \paragraph*{Background:} Assembly of short read sequencing data can result
  in a fragmented series of contigs. Therefore a common step in a genome
  project is to join neighbouring sequence regions together and fill gaps
  using additional sequence. This scaffolding step, however, is non-trivial
  and requires manually editing large blocks of nucleotide sequence. Joining
  sequence contigs together also hides the source of each region in the final
  genome sequence. Taken together, these considerations may make reproducing
  or editing an existing genome build difficult.

@@

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple to write and edit
  scaffold file syntax. This syntax further allows  unknown regions to be
  defined, and inserts to be added to the scaffold to fill contig gaps.
  Defining the genome construction in a file makes the scaffolding process
  reproducible and comparatively easier to edit and compare than raw fasta.

  \paragraph*{Conclusions:} Scaffolder is easy to use genome scaffolding
  software. This tool promotes reproducibility and maintenance in building
  a genome. Scaffolder can be found at \scaffolder.

@@

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each read represents the machine-interpreted
nucleotide sequence from a short region in the genome. At the time of writing,
high throughput sequencing is limited to producing reads less than $>$1000
nucleotides long. Therefore a sequence longer than this length, e.g.
a complete genome, these numerous individual reads must be pieced together.
The process of joining reads into longer sequences is the `assembly' stage in
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence compose of
these individual reads. An analogy for this process is a jigsaw puzzle. Each
nucleotide read represents a single piece and the final genome sequence is the
completed puzzle. Repetitive nucleotide `repeat' regions within a genome or
biased and incomplete sequencing data may however prevent the genome being
assembled into a continuous sequence. This is analogous to missing jigsaw
pieces or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}. Current genome assembly software is however
often unable to produce a complete sequence from these reads due to these
problems. In many cases assembly algorithms will generate several large
contiguous regions of sequence (`contigs') composed of many individual reads.
These contigs represent the assembly of reads into the longest possible
sequences. These contigs represent a fragmented picture of the genome and
require additional work to piece together. These contigs represent
a fragmented picture of the genome and require additional work to piece
together. \pb

The process of ameliorating a draft sequence to a finished continuous sequence
can be expensive in time and laboratory effort. Therefore in many cases a set
of contigs may often describe enough of the genomic content for many research
questions \cite{branscomb2002}. A continuous high-quality `finished' genome
sequence does however provide further depth of information including complete
resolution of repeat regions and the exact distances between genomic elements
\cite{parkhill2002,fraser2002}. This process of joining contigs together to
form a continuous sequence may be called the `scaffolding' or `finishing'
stage. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous sequence. Due to genomic complexity and missing data,
scaffolding may not ultimately produce a final completed sequence. The process
of scaffolding may still however succeed in joining a subset of contigs
together or using additional regions of sequence to fill gaps. An overview of
the scaffolding process is outlined in the following section.

@@

The sequencing process produces reads from either strand of the DNA helix.
Therefore the resulting contigs constructed from these reads may point in
either direction of the DNA strand. Contig orientation reverse complements
sequences where necessary so that all contigs point in the same direction and
represent the same DNA strand. In the case of microbes this orientation may be
the 5'-3' strand leading away from the origin of replication.

@@

Contig ordering determines the placement of contigs to best represent their
order in the true genome sequence. Correctly placing each contig may also
highlight any extra-genomic DNA, such as plasmids, and which should be treated
separately from genomic sequence. The contig order is commonly referenced
against the origin of replication where the contig containing the replication
start point is first. All subsequent contigs are then ordered in the direction
of DNA replication.

@@

Given the order and orientation of contigs, determining the correct distances
between contigs leads to an estimate of the genome size. The size of inter
contig gaps represents the length of an unknown region in the genome. Filling
in these regions in with unknown nucleotide characters (`N') allows a draft
continuous representation of the genome. This draft scaffold sequence is
useful for representing both the sequenced and unknown regions in the genome.

@@

During the scaffolding process closing and filling sequence gaps improves the
accuracy of the genome scaffold. This may require returning to the laboratory
to perform additional sequencing or using computational methods of estimating
the gap sequence. These additional sequences can be used to bridge and close
the gaps between contigs. Once all contigs have been joined and gaps in
a scaffold closed the genome may be considered finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both.
An example of a computational method uses pair read data, where the sequencing
reads are a known distance apart in the genome, when paired reads are in
separate contigs this can to estimate the order and distance between them.
Laboratory methods may use PCR to amplify the unknown DNA in gap regions,
followed by traditional Sanger sequencing. Computation methods are more
preferable as they less costly in laboratory time and materials compared to
manual gap resolution \cite{nagarajan2010}. The need for computational
scaffolding methods has resulted in a variety of algorithmic approaches. \pb

Synteny methods use sequence similarity to compare the assembled contigs to
a more complete reference genome sequence. Areas of corresponding sequence
between the the two sets are indicative of the possible contig organisation in
the true genome \cite{richter2007,zhao2008}. \pb

Repeat regions may be responsible for many of the gaps in building a genome
sequence. The existence of tandemly repeated nucleotide regions produces many
reads of similar sequence. The assembly of these regions may result in the
repeats being collapsed into a single instance or ignored by conservative
assembly algorithms. These regions may be reassembled using algorithms
specifically for repeat resolution \cite{mulyukov2002,koren2010}. A related
approach uses uniquely matching reads to try and bridge across contig gaps
between contigs \cite{tsai2010}. \pb

As mentioned above, the genome may be prepared prior to sequencing so that
reads are produced in pairs at a known distance. This paired-read data
provides an extra level of information on how contigs may be scaffolded
together. Heuristic graph algorithms may take advantage of this data to search
for optimal configuration of contigs in the scaffold to take this into account
\cite{dayarian2010}. Additional data using a reference genome may also be
combined when estimating the best contig configuration \cite{pop2004}. \pb

@@

Taken together these methods provide a wide variety of possible approaches for
scaffolding a genome. The scaffold may however still require manual editing
and addition of sequence if computational methods are not completely
successful. This process involves moving and joining larges blocks of
nucleotide text together. This process may be expected to be error-prone and
precludes any reproducibility of the scaffold. \pb

The software outlined here ``Scaffolder'' is aimed at creating a syntax and
software framework for producing and editing a genome scaffold. Scaffolder
uses a specific data syntax to define how contigs are joined, additional
sequences inserted, and for the specification of unknown regions. This syntax
allows a scaffold to be edited in a simpler way and allows the scaffolding
process to be reproducible and deterministic. \pb

@@

Unit tests were implemented to develop and maintain the individual elements of
the source code.  Integration tests were written to test the Scaffolder
software interface as a whole. Unit tests were written using the Shoulda and
RSpec \cite{rspec} libraries.  Integration tests were written using the
Cucumber library \cite{rspec}. \pb

The Scaffolder source code is documented using the Yard library \cite{yard}.
Unix manual pages for the command line interface were generated using the Ronn
library \cite{ronn}. The manipulation of biological sequences in Scaffolder is
performed using the BioRuby library \cite{goto2010}. A full list of the
software dependencies in Scaffolder can be found in the Gemfile in the root of
each source code directory. \pb

\subsection*{Scaffold file syntax} %{{{2

The choice of nucleotide sequences comprising the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a language format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML parser and generator
exists. The YAML website lists current parsers for languages including C/C++,
Ruby, Python, Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being
widely supported YAML formatted scaffold files can be validated for correct
syntax using third party tools such as Kwalify \cite{kwalify}. \pb

As described in the Background section, the initial sequencing data assembly
may result in an incomplete genome build. The use of computational software and
generation of additional sequence using PCR translates into genome scaffolding
as an on-going process. Therefore the scaffold file should be simple to edit
update manually in addition to being computationally tractable. This
requirement was therefore the second reason YAML was selected, where YAML is
human-readable and should be relatively easy to manipulate in a standard text
editor. \pb

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence, most likely a contig, in the genome scaffold. The
Scaffolder software processes this list of entries in order, converting each to
the corresponding nucleotide sequence. These nucleotide sequences are then
concatenated together to produce the draft super sequence. Each entry in the
scaffold file may have attributes that define whether a sub-sequence or the
reverse complement of the sequence should be used. The types of attributes
available and an example scaffold file are outlined in the Results and
Discussion section below.  \pb

@@

The Scaffolder software outlined here facilitates or computational joining of
nucleotide sequences together into a single contiguous super sequence.
Plain-text scaffold files written in the YAML syntax are used to specify how
nucleotide sequences should be joined together. This scaffold file is read by
the software to determine how the draft scaffold sequence should be generated.
In addition to specifying which contigs should be used, the specification of
the scaffold file allows the contigs to trimmed down to smaller sub-sequences
and reverse complemented if necessary. \pb

The process of genome finishing may involve producing additional
oligonucleotide sequence to fill areas of unknown sequence in the scaffold.
The scaffold file format allows these additional insert sequences to be added
and bridge gaps. These inserts can also be treated in a similar way to larger
contig sequences: trimmed and/or reverse complemented to match the
corresponding gap region. \pb

In the process of building a genome scaffold distances between contigs may be
estimated from paired read data or from mapping the contigs to a reference
genome. These inter contig gap regions are useful to join regions of sequence
together by an estimated size. The scaffold file also allows for the
specification of these unknown regions to highlight areas in the genome
sequence still to be resolved. \pb

The scaffold file is used to specify how sequence regions are joined together,
however the sequences themselves are not stored in the scaffold file. The
sequences are instead maintained in a fasta file. A fasta file of contigs and
scaffolded contigs is a common output for assemblers and as such should be
readily available without further processing required. Nucleotide sequence are
reverenced in the scaffold using the first space-delimited word from each fasta
header. Maintaining the nucleotide sequences in a separate file preserves the
original sequence and furthermore decouples the raw sequence from the markup
describing how it should be used to generate the genome sequence. \pb

\subsection*{Defining a Scaffold in a Text File} %{{{2

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a scaffold and the corresponding genome build. The basic
layout of the scaffold file is a list - a series of entries where each entry
corresponds to a region of sequence in the scaffold super sequence. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. The first
entry in the scaffold begins on Line 2. The first line of this entry,
highlighted in green, begins with a dash character `-', dashes are used to
denote entries in a YAML list. This first line therefore indicates indicate
the first entry in the scaffold. The first line of each entry in the scaffold
begins with a dash line. The second line of this entry indented by two spaces
where whitespace is used to group similar attributes together. Lines indented
by whitespace are linked to the first non-indented preceding line in the file.
\pb

The ``sequence'' tag indicates this entry corresponds to a sequence in the
fasta file. The following line indicates the name of the fasta sequence using
the ``source'' tag. As described above the first space-delimited word of the
fasta header is used to identify which sequence. Therefore these three lines
describe the first entry in the scaffold as a simple sequence using the fasta
sequence named `sequence1'. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line in the entry specifies the size of this unknown
region. This entry therefore inserts a region of 20 `N' characters in the
scaffold. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the fasta entry
`sequence2' to the scaffold. This entry demonstrates how the nucleotide
sequence may be manipulated prior to addition to the scaffold. The `start' and
`stop' tags trim the sequence to these coordinates inclusively. The
``reverse'' tag reverse complements the sequence. \pb

This region further includes the ``inserts'' tag to update this sequence with
regions of sequence. These inserts are also added as a YAML list with each
insert beginning with a dash line. The first insert, shown in purple,
illustrates using similar attributes to that of a sequence entry. The reverse,
start and stop tags, are used to trim and reverse complement the insert.
Similarly the `source' tag identifies the corresponding fasta sequence. The
``open'' and ``close'' tags determine where the insert is added in the
enclosing sequence. These coordinates inclusively replace the sequence region
within these coordinates with the specified insert fasta sequence. \pb

The next insert, shown in brown, specifies only the open position tag. This
illustrates that only one of either `open' or `close' tags is required. When
only one of these tag is used the corresponding opposing coordinate position
is calculated from the length in the insert fasta sequence. This allows
inserts to bridge into, and partially fill gap regions. \pb

@@

The sequence entry in the scaffold file allows multiple sequence-editing
operations to be applied. If applied in different combinations, these
operations could result in different output sequence. An example would be
reverse complementing the sequence then trimming the start of the sequence by
five nucleotides. Performing these actions in the reverse order would produce
a difference output sequence. This section therefore outlines the Scaffolder
protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts, these inserts are reverse sorted by the insert close coordinate. The
insert with the largest close position is first in the list and the insert
with the smallest close position is last. This preserves the original intended
coordinates for each insert during the process of adding each insert to the
sequence. \pb

@@

If the insert added to the sequence is a different size to that of the
replaced region then the sequence start and stop co-ordinates are updated. For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position order the remaining
inserts are unaffected by changes in the encoding sequence size. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
manipulation of the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. \pb

@@

Scaffolder relies on both the Ruby programming language and the RubyGems
package management system. These are either already present or easily
installed on recent distributions of Mac OSX or Linux. The presence of these
tools can be tested by typing the following on the command line. \pb

@@

If both these commands return without error this indicates the necessary
components are available to install and run Scaffolder. The Scaffolder API and
CLI can then be installed using the following command. \pb

@@

To ensure that the Scaffolder CLI is accessible, the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
\~/.gem this can be done by adding the following to the \~/.bashrc file. The
corresponding login script should be updated for other shells such as Zsh,
etc.. \pb

@@

The Scaffolder CLI provides functions to validate the scaffold file and build
the draft scaffold sequence. Assuming Scaffolder has been installed and the
binary is available in the command line path the following command should run
without error. This will list the available Scaffolder functions. \pb

@@

Scaffolder provides a function to test the scaffold for overlapping insert
coordinates. Overlapping inserts will result in unexpected behaviour when
generating the scaffold sequence as the coordinates for the first insert may
change the insert location for subsequent inserts. The ``validate'' command
will check the scaffold for any such inserts and return them as a YAML list.
\pb

@@

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the Scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described in the CLI section
above. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable array and the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of
the classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class on the Scaffolder website. \pb

@@

Scaffolder is designed to be as simple to use as possible. Assuming the
existence of Ruby and RubyGems, Scaffolder can be installed in a single
command line step. This should negate any possible barriers from requiring
manual compilation of the source code. \pb

Scaffolder uses a minimal and compact syntax to described the genome scaffold.
This syntax is simple to construct and edit whilst being succinct and
readable. The goal of this syntax is to make genome scaffolds easy to
construct a common text editor. Furthermore Scaffolder provides extensive
documentation both for the CLI and for the Ruby API. Therefore should any part
of the scaffolding process be unclear the documentation should make an
solution easy to find. \pb

@@

Constructing a genome sequence using a text file means the generation the
scaffold super sequence both reproducible and deterministic. In comparison
joining large nucleotide sequences by hand cannot be reliably reproduced. The
layout of the scaffold file furthermore provides a human readable record of
how the sequence was constructed and decouples the layout of the contigs in
the scaffold from their source sequence. The same cannot be said for large
block of nucleotide sequences comprised of many source contigs. \pb

Configuring the final sequence in the scaffold file means the build is easier
to edit once already constructed. The YAML text format also allows comparison
of differences between scaffold builds using standard text comparison tools
such as diff. This makes scaffold files easier to store in version control
systems and affords genome finishers similar tools used in software
development. \pb

>>>>

% Scaffolder url
\urldef{\scaffolder}\url{http://next.gs}

@@

\author{
  Michael D Barton$^{1}$\correspondingauthor%
  \email{Michael D Barton\correspondingauthor  - mail@nucleotid.es}%
\and
  Hazel A Barton$^1$%
  \email{Hazel A Barton - bartonh@nku.edu}%
      }

@@

  \paragraph*{Background:} Assembly of short read sequencing data can result in
  a fragmented series of contigs. Therefore a common step in a genome project
  is to join neighbouring sequence regions together and fill gaps in the
  assembly using additional sequence. This scaffolding step, however, is
  non-trivial and requires manually editing large blocks of nucleotide
  sequence. Joining sequence contigs together also hides the source of each
  region in the final genome sequence. Taken together, these considerations may
  make reproducing or editing an existing genome build difficult.

@@

  \paragraph*{Results:} This software allows a genome build to be defined in
  terms of the constituent contigs using a relatively simple to write and edit
  scaffold file syntax. This syntax further allows  unknown regions to be
  defined, and inserts to be added to the scaffold to fill gaps. Defining the
  genome construction in a file makes the scaffolding process reproducible and
  comparatively easier to edit and compare than fasta sequence.

  \paragraph*{Conclusions:} Scaffolder is easy to use genome scaffolding
  software. This tool promotes reproducibility and continuous development in a
  genome project. Scaffolder can be found at \scaffolder.

@@

High throughput sequencing can produce hundreds of thousands to millions of
sequence reads from a genome. Each sequence represents the machine-read
nucleotides from a short region in the genome. At the time of writing, high
throughput sequencing is limited to producing reads less than $>$1000
nucleotides in length. Therefore to determine a sequence longer than this, such
as a complete genome, these numerous individual reads must be pieced together.
The process of joining reads into longer sequences is the `assembly' stage in
a genome project \cite{miller2010}. \pb

Assembly software takes the nucleotide reads produced by the sequencing
hardware and, in the ideal case, outputs a complete genome sequence compose of
these individual reads. An analogy for this process is a jigsaw puzzle. Each
nucleotide read represents a single piece and the final genome sequence is the
completed puzzle. Sequences of repetitive nucleotide `repeat' regions within
the genome, or biased and incomplete sequencing data may however prevent the
genome being assembled into a continuous sequence. This is analogous to missing
jigsaw pieces or pieces that fit to multiple other pieces. \pb

The recent advent of these current high throughput sequencing methods has lead
to a renewed interest into algorithms for solving the problem of genome
assembly \cite{pop2008,pop2009}. Genome assembly software may however be unable
to produce a complete sequence from these reads. In many cases assembly
algorithms will generate several large contiguous regions of sequence
(`contigs') composed of many individual reads given the available data and
assembly algorithm. These contigs represent the assembly of reads into the
longest possible sequences. These contigs represent a fragmented picture of the
genome and require additional work to join together into larger sequence
lengths. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in both time and laboratory effort. Therefore the
genomic data present in a set of contigs may often be enough for many research
questions \cite{branscomb2002}. A continuous high-quality `finished' genome
sequence does however provide a greater depth of information, such as complete
resolution of repeat regions and more precise estimates of distances between
genomic elements \cite{parkhill2002,fraser2002}. This process of joining
contigs together to form a continuous sequence may be called the `scaffolding'
or `finishing' stage. \pb

@@

Scaffolding is the process of joining a series disconnected contigs into
complete continuous sequence. Due to genomic complexity and missing data,
scaffolding may not ultimately produce a final completed sequence, but may
still succeed in joining a subset of contigs together or filling gaps. An
overview of the scaffolding process is outlined in the following section.

@@

The sequencing process generates reads from either strand of the DNA helix.
Therefore the resulting contigs constructed from these reads may point in
either direction of the DNA strand .The orientation of contigs to point in the
same direction and represent the same DNA strand requires reverse complementing
sequences where necessary. In the case of microbes this orientation may be the
5'-3' strand leading away from the origin of replication.

@@

Contig ordering determines the placement of observed contigs to best represent
their order in the true genome sequence. Correctly placing each contig also
highlights any extra-genomic DNA, such as plasmids, and should be scaffolded
separately from genomic sequence. The contig order is commonly referenced
against the origin of replication where the containing contig is split at the
replication start point. All subsequent contigs are then ordered in the
direction of DNA replication.

@@

Given the correct order and orientation, determining the distances between
contigs results in an estimate of the complete genome size. The size of inter
contig gaps represents the length of an unknown region in the genome. Filling
these regions with unknown nucleotide characters (`N') allows a draft
continuous representation of the genome. This draft scaffold sequence is useful
for representing both the sequenced and unknown regions in the genome.

@@

During the scaffolding process closing and filling gaps between contigs
improves the accuracy of the genome scaffold. This may require returning to the
laboratory to perform additional sequencing or using computational methods of
estimating the gap sequence. This additional sequence is used to replace the
gap between two contigs joining them into a single sequence if they are not
already scaffolded together. Once all contigs have been joined and gaps in
a scaffold closed the genome may be considered finished.

@@

The process of resolving a genome sequence and building a scaffold uses
\emph{in vitro} methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use paired-read data, where the
sequencing reads were generated at a known distance apart in the genome. The
occurrence of a pair of reds in separate contigs this can be used to estimate
the order and distance. Laboratory methods may use PCR to amplify the unknown
DNA in a gap region, then use traditional Sanger sequencing to determine the
sequence. Computation methods are more preferable as they are less costly in
laboratory time and materials compared to manual gap resolution
\cite{nagarajan2010}. This need for computational scaffolding methods has
resulted in a variety of algorithmic approaches. \pb

Synteny methods compare the assembled contigs to a more complete reference
genome sequence searching for ares of sequence similarity. Areas of
corresponding sequence between the the two sets are can be used to infer contig
organisation \cite{richter2007,zhao2008}. Recombination between the two genomes
can however reduce the effectiveness of this analysis. \pb

Repeat regions may be responsible for many of the gaps in building a genome
sequence. The existence of tandemly repeated nucleotide regions produces many
reads of similar sequence. The assembly of reads generated from repeats can
result in the assemble collapsing into a single repeat instance or ignored by
conservative assembly algorithms. These regions may be resolved by using
reassembly algorithms specifically for repeat resolution
\cite{mulyukov2002,koren2010}. Similarly a related approach may use uniquely
matching reads to try and bridge across gaps between contigs \cite{tsai2010}.
\pb

As mentioned above, the genome may be prepared prior to sequencing so that
reads are produced in pairs at a known distance. This paired-read data
provides an extra level of information on how contigs may be scaffolded
together. Heuristic graph algorithms may take advantage of this data to search
for optimal configuration of contigs in the scaffold using paired-read distances
\cite{dayarian2010}. Additional data using a reference genome may also be
combined when estimating the best contig configuration \cite{pop2004}. \pb

@@

Taken together these methods provide a wide variety of possible approaches for
scaffolding a genome. The scaffold may still however require manual editing and
addition of sequence if computational methods are not completely successful.
This involves moving and joining larges blocks of nucleotide text together by
hand which may be expected to be error-prone and precludes any reproducibility.
\pb

The software outlined here ``Scaffolder'' is aimed at creating a syntax and
software framework for producing and editing a genome scaffold. Scaffolder uses
a specific file format to define how contigs are joined, additional sequences
inserted, and for the specification of unknown regions. This syntax allows
a scaffold to be edited in a simpler way and allows the scaffolding process to
be both reproducible and deterministic. \pb

@@

Unit tests were implemented to maintain individual elements of the source code
during development. Integration tests were written to test the Scaffolder
software interface as a whole. Unit tests were written using the Shoulda and
RSpec \cite{rspec} libraries.  Integration tests were written using the
Cucumber library \cite{rspec}. \pb

The Scaffolder source code is documented using the Yard library \cite{yard}.
Unix manual pages for the command line were generated using the Ronn library
\cite{ronn}. The manipulation of biological sequences in Scaffolder is
performed using the BioRuby library \cite{goto2010}. A full list of the
software dependencies in Scaffolder can be found in the Gemfile in the root of
each source code directory. \pb

\subsection*{Scaffold File Syntax} %{{{2

The choice of nucleotide sequences comprising the scaffold is specified using
the YAML syntax \cite{yaml}. YAML is a language format using whitespace and
indentation to produce a machine readable structure. As YAML is a standardised
data format, third-party developers have the option to generate a genome
scaffold using any programming language for which a YAML library exists. The
YAML website lists current parsers for languages including C/C++, Ruby, Python,
Java, Perl, C\#/.NET, PHP, and JavaScript. In addition to being widely
supported YAML formatted scaffold files can be validated for correct syntax
using third party tools such as Kwalify \cite{kwalify}. \pb

As described in the Background section, the initial sequencing data assembly
may result in an incomplete genome build. The use of computational software and
generation of additional sequence using PCR translates into genome scaffolding
as an on-going process. Therefore the scaffold file should be simple to update
manually in addition to being computationally tractable. This requirement was
therefore the second reason YAML was selected, where YAML is human-readable and
easy to edit in a standard text editor. \pb

The scaffold file takes the form of a list of entries. Each entry corresponds
to a region of sequence, most likely a contig or previously scaffolded set of
contigs. The Scaffolder software processes this list in order, converting each
to the corresponding nucleotide sequence. These nucleotide sequences are then
concatenated to produce a draft super sequence. Each entry in the scaffold file
may have attributes that define whether a sub-sequence or the reverse
complement of the sequence should be used. The types of attributes available
and an example scaffold file are outlined in the Results and Discussion section
below.  \pb

@@

The Scaffolder software outlined here allows computational joining of
nucleotide sequences together into a single contiguous super sequence.
Plain-text scaffold files written in the YAML syntax specify how these
sequences should be joined. This scaffold file is read by the scaffolder
software and used to generate the scaffold sequence. In addition to specifying
which contigs, the scaffold file allows the contigs to be edited to produce
smaller sub-sequences or reverse complemented if necessary. \pb

The process of genome finishing may involve producing additional
oligonucleotide sequence to fill areas of unknown regions in the scaffold. The
scaffold file format allows these additional insert sequences to be added to
bridge these gaps. These inserts can also be treated in a similar way to larger
contig sequences: trimmed and/or reverse complemented to match the
corresponding gap region. \pb

Distances between contigs may be estimated from paired read data or from
mapping the contigs to a reference genome. These inter-contig gap regions are
useful to join regions of sequence together by an estimated size. The scaffold
file allows for the specification of these unknown regions to join contigs
together. The use of these regions in the scaffold highlights areas still to be
resolved. \pb

The scaffold file is used to specify how sequence regions are joined together
but the sequences themselves are not stored in the scaffold file. The sequences
are instead maintained as a separate fasta file. A fasta file of contigs and
scaffolded contigs is a common output for assemblers and as such should be
readily available. The nucleotide sequences in the fasta file are referenced in
the scaffold using the first space-delimited word from each fasta header.
Maintaining the nucleotide sequences in a separate file preserves the original
sequence and furthermore decouples the sequence from the markup describing how
it should be used to generate the genome sequence. \pb

\subsection*{Defining a Scaffold as a Text File} %{{{2

The scaffold file is written using the YAML syntax. YAML is a data format
designed to be both machine readable and easy to edit manually. An example
scaffold file is shown in Figure 1. This file illustrates the text attributes
used to describe a scaffold and the joining of sequences in the genome build.
The basic layout of the scaffold file is a list - a series of entries where
each entry corresponds to a region of sequence in the generated scaffold super
sequence. The right hand side indicates hows the entries are joined together.
This figure is for illustrative purposes and not to scale. \pb

@@

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. The first
entry in the scaffold begins on the second line. The first line of this entry,
highlighted in green, begins with a dash character `-', dashes are used to
denote entries in a YAML list. The first line of each entry in the scaffold
therefore begins with a dash line. The next line is indented by two spaces
- whitespace is required to group similar attributes together. Lines indented
by whitespace are part to the first non-indented preceding line. \pb

The ``sequence'' tag indicates this entry corresponds to a sequence in the
fasta file. The following line indicates the name of the fasta sequence using
the ``source'' tag. As described above the first space-delimited word of the
fasta header is used to identify which sequence is selected from the fasta
file. Together these three lines describe the first entry in the scaffold as
a simple sequence using the fasta sequence named `sequence1'. On the right hand
side of Figure 1 this produces the first region in the scaffold, also shown in
green. \pb

@@

The second entry in the scaffold, highlighted in orange, is identified by the
``unresolved'' tag, indicating a region of unknown sequence but known length.
The line below specifies the size of this unknown region. This entry therefore
inserts a region of 20 `N' characters in the scaffold. \pb

@@

The last entry in the scaffold, highlighted in blue, adds the fasta entry
`sequence2' to the scaffold. This entry demonstrates how the nucleotide
sequence may be manipulated prior to addition to the scaffold. The `start' and
`stop' tags trim the sequence to these coordinates inclusively. The ``reverse''
tag instructs scaffolder to also reverse complement the sequence. On the right
hand side of the sequence of Figure 1 this completes the scaffold sequence. \pb

This entry includes the ``inserts'' tag to update this sequence with regions of
sequence. These inserts are also added as a YAML list with each insert
beginning with a dash line. The first insert, shown in purple, illustrates
using similar attributes to that of a sequence entry. The reverse, start and
stop tags, are used to trim and reverse complement the insert. Similarly the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags are specific to inserts and determine where the insert is added
in the enclosing sequence. The region of the sequence inside these coordinates
inclusively is replaced with the specified insert fasta sequence. This is
highlighted in the putative scaffold in Figure 1 by the black lines bisecting
the blue sequence. \pb

The next insert, shown in brown, is specified using only the open position tag.
This illustrates that only one of either `open' or `close' tags is required.
When only one of these tag is used the corresponding opposing coordinate
position is calculated from the length in the insert fasta sequence. This
allows inserts to bridge into, and partially fill gap regions with requiring an
end co-ordinate position. \pb

@@

The the scaffold file format allows multiple sequence-editing operations to be
applied to a sequence or insert. If applied in different combinations, these
operations will produce different output sequences. An example would be reverse
complementing the sequence then trimming the start of the sequence by five
nucleotides. Performing these actions in the reverse order would result in five
nucleotides being removed from the opposite end. This section therefore
outlines the Scaffolder protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts, these inserts are reverse sorted by the insert close coordinate. The
insert with the largest close position is first in the list and the insert
with the smallest close position is last. This preserves the original intended
coordinates for each insert during the process of adding inserts to the
sequence. \pb

@@

If the insert added to the sequence is a different size to that of the
replaced region then the sequence start and stop co-ordinates are updated. For
example if a 7bp insert replaced a 5bp region in the sequence the stop
coordinate of the enclosing sequence would be incremented by the 2bp
difference. As inserts are added by reverse close position the remaining
inserts are unaffected by changes in the encoding sequence size. \pb

@@

Scaffolder provides a standardised set of Ruby classes and methods (API) for
interacting with the scaffold. This allows Scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as Rake. \pb

@@

Scaffolder relies on both the Ruby programming language and the RubyGems
package management system. These are either already present or easily installed
on recent distributions of Mac OS X or Linux. The presence of both these tools
can be tested by typing the following on the command line. \pb

@@

If these commands return without error this indicates the necessary components
are available to install and run Scaffolder. The Scaffolder API and CLI can
then be installed using the following RubyGems command. \pb

@@

To ensure that the Scaffolder CLI is accessible, the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed to
$\sim$/.gem this can be done by adding the following to the $\sim$/.bashrc
file. The corresponding login script should be updated for other shells such as
Zsh, etc.. \pb

@@

The Scaffolder CLI provides functions to validate the scaffold file and build
a draft super sequence. Assuming Scaffolder has been installed and the binary
is available in the command line path the following should run without error.
This command will list the available Scaffolder functions. \pb

@@

Scaffolder provides a function to test the scaffold for overlapping insert
coordinates. Overlapping inserts will result in unexpected behaviour when
generating the scaffold sequence as the coordinates for the first insert may
change the intended coordinates for subsequent inserts. The ``validate''
command will check the scaffold for any such inserts and return them as a YAML
list. \pb

@@

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the Scaffolder website (\scaffolder). The
following code snippet outlines how the API may be used to create a script to
replicate the function of the `scaffolder sequence' command described in the
CLI section above. \pb

@@

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable array and the penultimate line illustrates
the trivial code required to generate the scaffold sequence. Further details of
the classes and methods related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class on the Scaffolder website. \pb

@@

Scaffolder is designed to be as simple to use as possible. Assuming the Ruby
and RubyGems software are present, Scaffolder can be installed in a single
command line step. This should negate any possible barriers from requiring
manual compilation of the source code. \pb

Scaffolder uses a minimal and compact syntax to described the genome scaffold.
This syntax is simple to construct and edit whilst being succinct and readable.
The goal of this syntax is to make genome scaffolds easy to construct in
a common text editor. Furthermore Scaffolder provides extensive documentation
both for the CLI and for the Ruby API. Therefore should any part of the
scaffolding process be unclear the documentation should make an solution easy
to find. \pb

@@

Constructing a genome sequence using a text file makes generating a scaffold
super sequence both reproducible and deterministic. In comparison joining large
nucleotide sequences by hand cannot be reliably reproduced. The layout of the
scaffold file furthermore provides a human readable record of how the scaffold
was constructed and decouples the layout of contigs from their source sequence.
\pb

Configuring the final sequence in the scaffold file means the build is easier
to edit once already constructed. The YAML text format also allows comparison
of differences between scaffold builds using standard Unix tools such as diff.
This makes scaffold files easier to store in version control systems and allows
genome finishers methods similar to that of software development. \pb
